---
title: 'Filling PDF forms with staplr: Under the hood'
author: Ogan Mancarci
date: '2019-05-14'
slug: filling-pdf-forms-with-staplr-under-the-hood
categories:
  - Code
tags:
  - R
draft: true
---
```{r setup, include=FALSE}
# knitr::opts_chunk$set(cache = FALSE)
# knitr::opts_knit$set(root.dir = here::here('static/files/staplr_pdfs'))
library(DiagrammeR)
library(pdftools)
library(magick)
showPDF = function(file, crop){
	tempFile = tempfile(fileext = '.png')
	pdf_convert(file,filenames = tempFile,verbose = FALSE)
	pdf = image_read(tempFile)
	pdf %>% image_crop(crop)

}
```

`staplr` is an R package, written by [Priyanga Dilini
Talagala](https://github.com/pridiltal). It is a wrapper for the
`[pdftk](https://www.pdflabs.com/tools/pdftk-the-pdf-toolkit/)` which is a pdf
manipulation toolkit. The package can flip files around, "staple" (hehe)
different pdf files together or separate them into pages. It can
also fill PDF forms. For instance if we have [this](/files/staplr_pdfs/simpleForm.pdf) file.

```{r,echo = FALSE,message=FALSE,out.width=300}
showPDF('files/staplr_pdfs/pdf1.pdf','300x250')
```

We can fill it by doing:

```{r formFill}
library(staplr)
fields = get_fields("files/staplr_pdfs/pdf1.pdf")
fields$TextField$value = "I am writing this to the PDF form"
fields$CheckBox$value = "Yes"
fields$OptionButton$value = 1
set_fields("files/staplr_pdfs/pdf1.pdf",
		   "files/staplr_pdfs/output.pdf",
		   fields)
```

And you will get [this](/files/staplr_pdfs/output.pdf) file with the fields field.

```{r,echo = FALSE,message=FALSE,out.width=300}
library(pdftools)
library(magick)


showPDF('files/staplr_pdfs/output.pdf','300x250')
```

This isn't exactly a new feature. I
[suggested](https://github.com/pridiltal/staplr/issues/2) the feature on late
February 2018, when the package was about a month old, pushed the [first working
version](https://github.com/pridiltal/staplr/pull/3) on March 2018 and by May it
was on CRAN. It worked pretty much the same way it worked now. However in
reality I underestimated how complex a PDF could be even as I was interacting with it through `pdftk` which is doing the heavy lifting. In the past year, I have been iteratively fixing one edge case after another. Many of those issues could have been avoided if I skimmed the [PDF file specification](https://www.adobe.com/content/dam/acom/en/devnet/pdf/pdfs/PDF32000_2008.pdf) earlier... This is a tale of those fixes.

## Prototype

I wrote an early version of a PDF filling function using on October 2017, to
fill [Dungeons and Dragons](https://oganm.com/shiny/printSheetApp) character sheets.
My brief search on how to fill pdf forms lead me to `pdftk`. The way it filled pdfs appeared to be fairly simple.

```{r diag1,echo=FALSE,out.height=110}
drawGraph = function(graph){
	graph %>% 
		generate_dot() %>% 
		gsub(pattern = 'neato',replacement = 'dot',x= .) %>%
		gsub(pattern = "graph \\[",'graph \\[rankdir = LR,\n',x = .) %>% 
		grViz()
}

graph = create_graph() %>% 
	add_node(label = 'PDF file') %>%  # 1
	add_node(label = 'pdftk\ngenerate_fdf', # 2
			 node_aes = node_aes(width = .8)) %>% 
	add_node(label = 'FDF file') %>%  # 3
	add_node(label = 'pdftk\nfill_form') %>%  # 4
	add_node(label = 'Filled PDF', node_aes = list(width = 0.6)) %>%  # 5
	add_node(label = 'Edit FDF\nto fill fields', node_aes = list(width = 0.8)) %>% # 6
	add_node(label = 'Filled FDF', node_aes = list(width = 0.6)) %>% # 7 
	set_node_attrs('shape',
				   'rectangle',
				   nodes = c(2,4,6)) %>% 
	set_node_attrs('fillcolor',
				   'lightgray',
				   nodes = c(2,4,6)) %>%
	set_node_attrs(
		node_attr = 'fontcolor',
		values = 'black'
	)  %>%
	add_edge(from=1,to=2) %>%
	add_edge(from = 2, to = 3) %>%
	add_edge(from  = 4, to = 5) %>%
	add_edge(from =1, to =4) %>% 
	add_edge(from = 3, to = 6) %>% 
	add_edge(from = 6, to = 7) %>% 
	add_edge(from = 7, to = 4)

drawGraph(graph)

```

As shown in the diagram above, `pdftk` fill pdf files using an intermediate text file in the
FDF format. FDF format is described in the PDF specification (which I didn't read at the time)
but just looking at the file creates a nice illusion of simplicity. 

```{bash}
pdftk files/staplr_pdfs/output.pdf generate_fdf output files/staplr_pdfs/output.fdf
cat -v files/staplr_pdfs/output.fdf # have to use -v or knitr complains about the second line
```

This is the FDF output from the file we just filled and things look quite obvious.
You have field contents written as `/V` followed by field names written as `/T (FieldName)`. 
Text fields are written within `()` while button states are written after a `/` (in this case `/Yes` for the checked box). 

My original implementation simply read this file and replaced the relevant `/V` lines
with the correct text. 
```{r}
fdfEdit = function(x, field,fdf){
    if(x == TRUE & is.logical(x)){
        x = '/Yes'
    } else if (x == FALSE & is.logical(x)){
        x = '/Off'
    } else {
        x %<>%
    		# note the excessive use of escapes due to use of str_replace later
            gsub(x = ., pattern = '\\', replacement = "\\\\\\\\" , fixed=  TRUE) %>%
            gsub(x = ., pattern = '(',replacement = '\\\\(',fixed = TRUE) %>%
            gsub(x = ., pattern = ')',replacement = '\\\\)', fixed = TRUE)

        x = paste0('(',x,')')
    }

    fdf = stringr::str_replace(string  = fdf,pattern = paste0('/V\\s.*\n/T\\s\\(',field,'\\)'),
                               replacement = paste0('/V ',x,'\n/T \\(',field,'\\)'))
    return(fdf)
}
```

In the code above `x` is the content of the field, `field` is the name and `fdf` is a
character containing the entire fdf file. So to fill a pdf file I would have to do something like


```{r}
system('pdftk files/staplr_pdfs/pdf1.pdf generate_fdf output files/staplr_pdfs/output2.fdf')

fdf = readLines('files/staplr_pdfs/output2.fdf') %>% paste(collapse = '\n')

# write to text field
fdf = fdfEdit('I am writing this to the PDF form','TextField',fdf)
# check to checkbox
fdf = fdfEdit(TRUE,'CheckBox',fdf)

write(fdf,'files/staplr_pdfs/output2.fdf')

system("pdftk files/staplr_pdfs/pdf1.pdf fill_form files/staplr_pdfs/output2.fdf output files/staplr_pdfs/output2.pdf")
```

```{r,echo = FALSE,message=FALSE,out.width=300}
showPDF('files/staplr_pdfs/output2.pdf','300x250')
```

## First "working" version

This was were I was when I started to work on staplr. The current version was
enough for my personal use but even just looking at the example above, one can
see that it is not enough. You may have noticed that the version above does not
have a way to select the option buttons. Also if we look at the FDF file of the unfilled
form:

```{bash}
pdftk files/staplr_pdfs/pdf1.pdf generate_fdf output files/staplr_pdfs/pdf1.fdf
cat -v files/staplr_pdfs/pdf1.fdf
```

You can see that there is no distinction in the way the checkbox and the option 
button is encoded. For someone to be able to use this function as it is, they would
have to experiment with the file to see how the values change when certain buttons are
checked or certain options are selected. Ideally we would have a way of providing
more information about the fields to the end user. Thankfully, `pdftk` has a separate
command that creates a file with the required information,

```{bash}
pdftk files/staplr_pdfs/pdf1.pdf dump_data_fields output files/staplr_pdfs/pdf1_data
cat files/staplr_pdfs/pdf1_data
```

This is great, we see fields' current values, their types and possible states that
a field can be in. Why isn't something like this used as input in the first place?
That is not for me to know. 

This brings us to the basic way of how `staplr` currently
works. `get_fields` function will parse this file and return an R list that
contains information about the properties of the field. After editing this object
`set_fields` is used to generate an `fdf` file using the original `fdfEdit` function.

```{r}
fields = get_fields('files/staplr_pdfs/pdf1.pdf')
fields

fields$TextField$value ='Some text'

set_fields(input_filepath = 'files/staplr_pdfs/pdf1.pdf',
		   output_filepath = 'files/staplr_pdfs/pdf2.pdf',
		   fields = fields)
```
This alters our original pipeline a little

```{r}
graph = create_graph() %>% 
	add_node(label = 'PDF file') %>%  # 1
	add_node(label = 'pdftk\ngenerate_fdf', # 2
			 node_aes = node_aes(width = .8)) %>% 
	add_node(label = 'FDF file') %>%  # 3
	add_node(label = 'pdftk\nfill_form') %>%  # 4
	add_node(label = 'Filled PDF', node_aes = list(width = 0.6)) %>%  # 5
	add_node(label = 'Edit FDF\nto fill fields', node_aes = list(width = 0.8)) %>% # 6
	add_node(label = 'Filled FDF', node_aes = list(width = 0.6)) %>% # 7 
	set_node_attrs('shape',
				   'rectangle',
				   nodes = c(2,4,6)) %>% 
	set_node_attrs('fillcolor',
				   'lightgray',
				   nodes = c(2,4,6)) %>%
	set_node_attrs(
		node_attr = 'fontcolor',
		values = 'black'
	)  %>%
	add_edge(from=1,to=2) %>%
	add_edge(from = 2, to = 3) %>%
	add_edge(from  = 4, to = 5) %>%
	add_edge(from =1, to =4) %>% 
	add_edge(from = 3, to = 6) %>% 
	add_edge(from = 6, to = 7) %>% 
	add_edge(from = 7, to = 4)


graph %>%
	add_node(label = 'pdftk \n dump_data_fields', node_aes = list(width = 1.2)) %>%  # 8
	add_node(label = 'data fields\nfile') %>% # 9
	add_node(label = 'pdftk \n dump_data_fields', node_aes = list(width = 1.2)) %>%  # 8
	set_node_attrs('shape',
				   'rectangle',
				   nodes = c(8)) %>% 
	set_node_attrs('fillcolor',
				   'lightgray',
				   nodes = c(8)) %>% 
	set_node_attrs(
		node_attr = 'fontcolor',
		values = 'black'
	) %>% 
	add_edge(from=1,to=8) %>% 
	add_edge(from = 8, to = 9) %>% 
	drawGraph()
```